Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-27 paper/paper.tex
latest - 27th

\documentclass[preprint,9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{lineno}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}

@@

\newcommand{\comment}[1]{{\color{Myblue}{#1}}}
\newcommand{\warning}[1]{{\color{Myred}{#1}}}
%\newcommand{\blankout}[1]{}

\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{\mbox{S}}   
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}
\linenumbers
% --- Author Metadata here ---
%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
%over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

While this idea has been shown to be effective for a
first-order {\em  eager} language~\cite{asati14lgc}, a straightforward  extension of the
technique   is  not   possible  for  lazy   languages,  where
heap-allocated objects  may include closures  (runtime representations
of unevaluated expressions). Since data  is made live by evaluation of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness  analysis  itself.    In  addition, laziness
also complicates garbage collection, since, 
 apart  from
evaluated   data,  we  need to extend   liveness-based  collection   to
closures.
Since a closure can  escape the scope in which it  was created, it has
to carry the liveness information  of its free variables. Moreover, as
execution  progresses and  possible  future uses  are eliminated,  the
liveness information is updated with more precise versions.  All this
requires significant  modification to  the earlier  garbage collection
scheme for eager languages.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage collectors in reclaiming more garbage.  The specific
contributions of this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
  functional language and  prove its correctness. The correctness
  proof involves setting up a non-standard semantics as a
  specification of liveness and then proving that the analysis is
  correct with respect to the specification.   
\item Our  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  regarding  whether   to  copy  a  cell   during  garbage  collection
  translates  to the  membership problem  of such  grammars.  In  this
  paper, we  have shown the  undecidability of the  membership problem
  for  the  class  of  grammars  that result  from  our  analysis.  To
  circumvent the  problem, we  over-approximate the  grammars generated
  by the analysis with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  demonstrate the effectiveness of liveness-based garbage collection.
\end{itemize}
To the  best of our
  knowledge, this is the first work on liveness-based garbage
  collection      for      lazy     languages.       All      previous
  attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
    asati14lgc, karkare06effectiveness} have targeted either
  imperative or  eager functional languages.

@@

The  semantics  of   expressions  (and  applications\footnote{In  most
  contexts, we  shall use the  term 'expression' and the  notation $e$
  for both expressions and applications.}) are given by transitions of
the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
Here \stk\ is a stack of continuation frames.  Each continuation frame
is of the form $(\rho, \ell,  e)$, signifying that the location $\ell$
has  to  be  updated  with  the  value  of  the  currently  evaluating
expression, and $e$ is to be evaluated next in the environment $\rho$.
Note  that  the  state  of  the  transition  system  consists  of  the
\emph{current evaluation  context}, given by  $\rho$ and $e$,  and the
\emph{suspended  evaluation  contexts}  (represented  as  continuation
frames) in the stack $S$.   \warning{The
3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
system will be denoted $\mathcal{S}$.}

The  initial  state  of  the  transition  system  described  above  is
$([\;]_\rho,\,   ([\;]_\rho,   \,   \rho(\ans),\,   (\print~\ans)):[\;]_{S},
[\;]_{H},\, (\mainpgm))$,  in which  $[\;]_\rho$ and $[\;]_H$  are the
empty environment and  the empty heap. The  initial stack consists
of  a single  continuation frame  in  which \ans\  is a  distinguished
variable  that  would   eventually  be  updated  with   the  value  of
(\mainpgm).  In  addition, \print\ is  a function modeling  a printing
mechanism---a   standard   run-time   support  assumption   for   lazy
languages---that prints  the value of  (\mainpgm).  The operator $:$
pushes elements on top of the stack.

@@

A variable is {\em live} if there  is a possibility of its value being
used in  future computations and  dead if  it is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
liveness of  references.  Using  $\acar$, $\acdr$ to  represent access
using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
reachable from a  variable can be represented by a  set of {\em access
  paths}     i.e.     prefix-closed     set     of    strings     from
$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
$\{\epsilon, \acar, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$,
then future  computations can only  refer up  to the second  and third
members  of $x$.   A  {\em  liveness environment}  is  a mapping  from
variables to access paths, but often expressed as
a set,  for example  by writing $\{x.\epsilon,  x.\acdr, x.\acdr\acdr,
y.\epsilon\}$  instead  of  $[x  \mapsto\{\epsilon,\acdr,\acdr\acdr\},
  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $y \mapsto
\{\epsilon\}$ represents access using $y$  itself and $z \mapsto \{\}$
indicates that $z$ is dead. 
In lazy languages, liveness environments are associated with regions of
programs instead of program points.

A notion that generalizes liveness is {\em demand}.  While liveness 
gives the patterns of future uses of a  variable, demand
represents the future  use of the value of an  expression.  The demand
on an expression  $e$ is again a set of  access paths---that subset of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the  let expression
$\pi:(\LET~x\leftarrow  (\CDR~y)~\IN~ \pi':\SRETURN~x)$.   Assume that
the context  of this expression  places the demand  $\lbrace \epsilon,
\acar \rbrace$. Since the value of the expression is
the value  of $x$, the demand  translates to the liveness  $[x \mapsto
  \lbrace   \epsilon,  \acar   \rbrace]$  at   $\pi'$.   Due   to  the
\LET\ definition which binds $x$ to $(\CDR~y)$, the liveness of $x$ at
$\pi'$ now becomes the demand on $(\CDR~y)$.  This, in turn, generates
the liveness  $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar  \rbrace$ at
$\pi$.  These are  the set of $y$-rooted accesses  required to explore
$\lbrace \epsilon, \acar  \rbrace$ paths of the  result of $(\CDR~y)$.
%% As an analogy  with classical (strong) liveness analysis,  $y$ and $z$
%% are live  at the entry $\pi:  x:=y+z$, if and  only if $x$ is  live at
%% exit   of  $\pi$.    In   our  terminology,   the  liveness   $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand on
%% $y+z$, and this,  in turn generates the  liveness $\lbrace y.\epsilon,
%% z.\epsilon \rbrace$ at the entry of $\pi$.

We use $\sigma$
to  range over  demands,  $\alpha$  to range  over
access  paths and  $\Lv$  to  range over  liveness
environments.   The   notation  $\sigma_1\sigma_2$
denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
\alpha_1     \in     \sigma_1,    \alpha_2     \in
\sigma_2\rbrace$.  Often  we shall  abuse notation
to juxtapose  an edge  label and  a set  of access
paths;   $\acar\sigma$   is    a   shorthand   for
$\lbrace\acar\rbrace\sigma$.

@@

\warning{Note  carefully that  $\Lv_y$ specifies  the liveness  of the
  reachable heap starting  from $y$ and extending  till closures.  The
  liveness  of further  heap  contents starting  from  a closure,  say
  $(\CAR\, x_{\pi})$, is given  by $\Lv_{x_{\pi}}$.  Therefore, in our
  implementation,   a  closure   carries   the  liveness   environment
  consisting of its free variables.}

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.  \warning{While  the proof  is on
  the broad lines of  \cite{asati14lgc}, there are several differences
  due to the  lazy semantics of the language in  this paper.} Here are
the ideas behind the proof.
\begin{enumerate}
\item       We        enrich       the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
  collection  just  before  each  $\rightsquigarrow$  transition.  The
  simulated garbage collection starts from  the root-set and inserts a
  special value  $\bot$ at each location  in the heap that  contains a
  reference that is reachable but not live. Any attempt to dereference
  such  locations during  the  transition will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

\hline
\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\end{figure*}

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item   We   enrich    the   abstract   machine   state
  $\rho,S,H,e$   to  $\rho,S,H,e,\sigma$. 
We  call such a state  a \emph{minefield state}.
Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
  $e$, i.e., it arises
  from the  actual sequence of function  calls that led
  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
  %% during    liveness     analysis    is     a    static
  %% over-approximation   of   $\sigma$   that   considers
  %% \emph{all  possible calls}  that  could  lead to  the
  %% evaluation of  $e$.
  The  information in  a suspended
  evaluation  context  on  the   stack  $S$  is  also
  similarly augmented with its  demand.  Thus a stacked
  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
  %% modification  of  the   small  step  semantics  which
  %% carries the extra information  is shown in the bottom
  %% table of Figure~\ref{fig:minefield-semantics}.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This represents the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate}
  \item Given the current  evaluation context $\rho, S,
    \heap,   e,  \sigma$   and   the  demand   transformers
    $\Lfonly$, we construct a liveness environment
    \Lv.  %%  If  $e$  is  an expression,  then  $  \Lv  =
    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
    %% application, then $ \Lv = ref(e, \sigma, \Lfonly)$.
    We  similarly  define the  liveness environment
    for each  of the  suspended evaluation  contexts in
    $S$, giving  a stack of liveness  environments that
    we shall denote $\mathsf{SL}$.
  \item \emph{Garbage collection  of objects pointed by
    root   set:}  For   each   $x  \in   domain(\rho)$,
    $\rho'(x)=\bot$     iff      $x.\epsilon     \notin
    \Lv$.    Similarly,    for   each    stack    entry
    $(\rho,\_,\_,\_)$  in   $S$  with  $\Lv'$   as  the
    corresponding      liveness     environment      in
    $\mathsf{SL}$, and  for each $x  \in domain(\rho)$,
    $\rho'(x)=\bot$ iff $x.\epsilon \notin \Lv'$.
  \item  \emph{Garbage  collection  of objects  in  the
    heap:} \warning{ For each location  $\ell$, $\heap'(\ell) = \bot$
    iff  there   is  no  $x$  in   either  the  current
    environment  or  one  of the  stacked  environments
    $\rho$, such  that for some  $\alpha$ corresponding
    to a  \warning{closure-free path}  in the heap  starting from
    $x$,  $\heap[x.\alpha]   =  \ell$  and   $x.\alpha  \in
    \Lv$}. Here \Lv\ is the liveness of the variables in
    the current environment or the stacked environment.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{H}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in  the  top   table  of
Figure~\ref{fig:minefield-semantics}.  If  this  or  an
earlier  call to  $GC$  results  in $H(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state {\sc  car-clo-bang}. Otherwise
the transition is the same as the earlier {\sc car-clo}
rule.

@@

\warning{We shall call a unit  allocatable memory as a {\em
  cell}.    A  cell   can  hold   a  basic   value
($\mathsf{bas}$),          the         constructor
\CONS\   $(\mathsf{cons~   arg_1~arg_2}$)   or   a
closure.   The closure,  in  turn, can  be one  of
($\mathsf{unop~arg}$),                ($\mathsf{
  binop~arg_1~arg_2}$)  and  function  application
($\mathsf{f~arg})$.  Here each $\mathsf{
  arg_i}$ is  a reference to another  heap cell.  In
addition,  the closure  also carries a pointer to
an automaton (denoted $\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.}

>>>>

\documentclass[preprint,9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}

@@

\newcommand{\comment}[1]{{\color{Myblue}{#1}}}
\newcommand{\warning}[1]{{\color{Myred}{#1}}}
%\newcommand{\blankout}[1]{}
\newcommand{\oxlb}{\ensuremath{[\!|}}
\newcommand{\oxrb}{\ensuremath{|\!]}}
\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{{S}}      
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}
\linenumbers 
% --- Author Metadata here ---
%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
%over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE. 
% --- End of Author Metadata ---

@@

While this idea has been shown  to be effective for a first-order {\em
  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
the technique is not possible for lazy languages, where heap-allocated
objects may  include closures (runtime representations  of unevaluated
expressions).   Firstly, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.  Secondly,  for liveness-based
garbage collection to  be effective, we need  to extend liveness-based
collection to closures apart from evaluated data.  Since a closure can
escape the scope in which it was created, it has to carry the liveness
information of  its free variables. Moreover,  as execution progresses
and  possible  future uses  are  eliminated,  we update  the  liveness
information  in a  closure with  a more  precise version.   Because of
these reasons,  the garbage collector also  becomes significantly more
complicated than a liveness-based collector for a eager language.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage collectors in reclaiming more garbage.  The specific
contributions of this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
  functional language and  prove its correctness. The correctness
  proof involves setting up a non-standard semantics as a
  specification of liveness and then proving that the analysis is
  correct with respect to the specification.   
\item Our  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  regarding  whether   to  copy  a  cell   during  garbage  collection
  translates  to the  membership problem  of such  grammars.  In  this
  paper, we  have shown the  undecidability of the  membership problem
  for  the  class  of  grammars  that result  from  our  analysis.  To
  circumvent the  problem, we  over-approximate the  grammars generated
  by the analysis with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  demonstrate the effectiveness of liveness-based garbage collection.
\end{itemize}
To the  best of our
  knowledge, this is the first work on liveness-based garbage
  collection      for      lazy     languages.       All      previous
  attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
    asati14lgc, karkare06effectiveness} have involved either
  imperative or  eager functional languages.

@@

The  semantics  of   expressions  (and  applications)
%% \footnote{In  most
%%   contexts, we  shall use the  term 'expression' and the  notation $e$
%%   for both expressions and applications.} 
are given by transitions of
the form $\rho, \stk, \heap,  e \rightarrow \rho', \stk', \heap', e'$.
Here \stk\ is a stack of continuation frames.  Each continuation frame
is of the form $(\rho, \ell,  e)$, signifying that the location $\ell$
has  to  be  updated  with  the  value  of  the  currently  evaluating
expression, and $e$ is to be evaluated next in the environment $\rho$.
Note  that  the  state  of  the  transition  system  consists  of  the
\emph{current evaluation  context}, given by  $\rho$ and $e$,  and the
\emph{suspended  evaluation  contexts}  (represented  as  continuation
frames) in the stack $S$.   
%% \warning{The
%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
%% system will be denoted $\mathcal{S}$.}

The  initial  state  of  the  transition  system  described  above  is
$([\;]_\rho,\,   (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
(\print~\ans)):[\;]_{S}, 
[\;]_{\heap},\, (\mainpgm))$,  in which  $[\;]_\rho$,
$[\;]_{\mbox \footnotesize {H}}$
and $[\;]_{S}$ are    are the
empty environment, heap and stack respectively. The  initial stack consists
of  a single  continuation frame  in  which \ans\  is a  distinguished
variable  that  would   eventually  be  updated  with   the  value  of
(\mainpgm) and $\rho_\mathit{init}$ maps \ans\ to a location $\ell_{\ans}$.   In  addition, \print\ is  a function modeling  a printing
mechanism---a   standard   run-time   support  assumption   for   lazy
languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The operator $:$
pushes elements on top of the stack.

@@

A variable is {\em live} if there  is a possibility of its value being
used in  future computations and  dead if  it is definitely  not used.
Heap-allocated data  needs a richer model  than classical liveness---a
model  which  talks  about  liveness of  references.   Using  $\acar$,
$\acdr$  to  represent access  using  $\CAR$  and $\CDR$  fields,  the
liveness of the structure reachable from a variable can be represented
by a set of {\em access paths} i.e.  prefix-closed set of strings from
$\{\acar,\acdr\}^\ast$.  Given an initial location $\ell$ (typically a
variable  reference) and  a heap  \heap, semantically  an access  path
$\alpha$ represents  a reference, denoted $\heap\oxlb\ell,  \alpha\oxrb$, in
the heap  that is  obtained by  starting with  $\ell$ and  chasing the
\CAR\ or  \CDR\ fields  as specified  by the  access path.  
$\heap\oxlb\ell, \alpha\oxrb$  is defined only  if the path followed  in the
heap  is \emph{closure-free}  (does not  cross closures),  else it  is
undefined.

Access paths can be used to  represent liveness. As an example, if $x$
is  a  list  with  liveness  $\{\epsilon,  \acar,  \acdr,  \acdr\acar,
\acdr\acdr,  \acdr\acdr\acar\}$,  then  future computations  can  only
refer up  to the  second and  third members of  $x$.  A  {\em liveness
  environment} is a mapping from  variables to access paths, but often
expressed as  a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
  \mapsto\{\epsilon,\acdr,\acdr\acdr\},          y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.    In  this   notation,  $y   \mapsto  \{\epsilon\}$
represents access using $y$ itself and $z \mapsto \{\}$ indicates that
$z$ is dead.  In lazy  languages, liveness environments are associated
with regions of programs instead of program points.

A notion  that generalizes liveness  is {\em demand}.   While liveness
gives the patterns of future uses of a variable, demand represents the
future use of the value of an expression.  The demand on an expression
$e$   is   again   a   set    of   access   paths---that   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the  let expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   \SRETURN~x)$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the  set  of $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.

We use $\sigma$  to range over demands, $\alpha$ to  range over access
paths  and $\Lv$  to range  over liveness  environments. $\Lv_y$  will
denote  the liveness  of an  individual  variable $y$  in $\Lv$.   The
notation $\sigma_1\sigma_2$ denotes  the set $\lbrace \alpha_1\alpha_2
\mid \alpha_1  \in \sigma_1, \alpha_2 \in  \sigma_2\rbrace$.  Often we
shall abuse  notation to juxtapose an  edge label and a  set of access
paths; $\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.

@@

As noted earlier, $x.\alpha \in  \Lv_y$ specifies the liveness  of the
reference $\heap\oxlb\rho(x), \alpha\oxrb$  only if $\alpha$ corresponds to  a closure-free path
in \heap\   starting from $\rho(x)$.  If this  path is intercepted  by a
closure,  say  $\CAR~{x_{\pi}}$, then  the  liveness  of further  heap
contents starting from $x_{\pi}$ is given by $\Lv_{x_{\pi}}$.

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.  %% \warning{While  the proof  is on
  %% the broad lines of  \cite{asati14lgc}, there are several differences
  %% due to the  lazy semantics of the language in  this paper.}
Here are
the ideas behind the proof.
\begin{enumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
  collection  just  before  each  $\rightsquigarrow$  transition.  The
  simulated garbage collection starts from  the root-set and inserts a
  special value  $\bot$ at each location  in the heap that  contains a
  reference that is reachable but not live. Any attempt to dereference
  such  locations during  the  transition will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

\hline
\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for \sc{car-clo-bang} and
  \sc{car-clo-bang} \label{fig:minefield-semantics-for-some}}
\end{figure*}

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item   We   extend    the   abstract   machine   state
  $\rho,S,H,e$   to  $\rho,S,H,e,\sigma$. 
We  call such a state  a \emph{minefield state}.
Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
  $e$, i.e., it arises
  from the  actual sequence of function  calls that led
  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
  %% during    liveness     analysis    is     a    static
  %% over-approximation   of   $\sigma$   that   considers
  %% \emph{all  possible calls}  that  could  lead to  the
  %% evaluation of  $e$.
  The  information in  a suspended
  evaluation  context  on  the   stack  $S$  is  also
  similarly augmented with its  demand.  Thus a stacked
  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
  %% modification  of  the   small  step  semantics  which
  %% carries the extra information  is shown in the bottom
  %% table of Figure~\ref{fig:minefield-semantics}.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This represents the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item \label{gc-1}Using the current  evaluation context $\rho, \stk,
    \heap,   e,  \sigma$   and   the  demand   transformers
    $\Lfonly$, we construct a liveness environment \Lv\ for the
    expression $e$.
      %%  If  $e$  is  an expression,  then  $  \Lv  =
    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
    %% application,  then  $  \Lv  =  ref(e,  \sigma,  \Lfonly)$.  
 We  similarly  define  the  liveness  environment  for  the
 expressions $e$ in each of the suspended evaluation contexts $(\rho, \ell,  e, \sigma)$ in $S$, giving
 a stack of liveness environments that we shall denote $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current  context and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff there  is no
    variable  $x$  in either  the  current  environment or  a  stacked
    environment,  say  $\rho$,  such   that  $x.\alpha$  is  live  and
    corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
    domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
    \oxrb = \ell$  and $x.\alpha \in \Lv$, where \Lv\  is the liveness
    environment corresponding to $\rho$ mentioned in step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{H}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $H(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
the transition is the same as the earlier {\sc car-clo}
rule.

@@

We shall call a  unit allocatable memory as a {\em  cell}.  A cell can
hold    a    basic    value    ($\mathsf{bas}$),    the    constructor
\CONS\ $(\mathsf{cons~  arg_1~arg_2}$) or a closure.   The closure, in
turn,    can    be    one   of    ($\mathsf{unop~arg}$),    ($\mathsf{
  binop~arg_1~arg_2}$)  and  function application  ($\mathsf{f~arg})$.
Here each $\mathsf{  arg_i}$ is a reference to another  heap cell.  In
addition, the closure also carries  a pointer to a DFA (denoted
$\mathsf{arg_i.dfa_i}$) for each $\mathsf{arg_i}$.
